---
title: "2.0.1_Improved"
author: "Theo Snow"
date: "2025-08-01"
output:
  pdf_document: default
  html_document: default
---
## **0. Package Calls, Options, and General Helper Functions**
```{r}
# ---- options ----
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  cache = FALSE,    
  fig.width = 8,
  fig.height = 5
)
options(scipen = 999)  # 

# ---- libraries ----
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(
  tidyverse, usmap, sf, broom, broom.mixed, scales, data.table,
  tableone, knitr, kableExtra, gt, gtsummary, patchwork,
  car, lmtest, sandwich, pscl, glmnet, conflicted, rgeoda, geodaData
)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)

# ---- paths helper ----
crypto_data <- read.csv("") %>%
  rename("Year" = "YEAR")
Weighted_denominators <- read.csv("")
state_abs <- data.frame(state.abb, state.name) %>%
  mutate(state.name = tolower(state.name))
Unweighted_denominators_month <- read.csv("") %>%
  rename("Year" = "date2_year", "Month" = "date2_month")
Unweighted_denominators_year <- read.csv("") %>%
  rename("Year" = "date2_year") %>%
  filter(Year >=2013 & Year <=2023)
climate_regions_map <- read.csv("us_climate_regions.csv") %>%
  rename("prefstate" = "state_code")


# ---- constants ----
ethrace_levels <- c("Hispanic","White", "African American", "Asian","Native American", "Other", "Unknown")
age_levels_table1 <- c("Under 5", "5 to 9","10 to 14", "16 to 19","20 to 24", "25 to 29", "30 to 34", "35 to 39",
                       "40 to 44", "45 to 49", "50 to 54", "55 to 59","60 to 64", "65 and Over", "Unknown")
gender_levels <- c("Male", "Female", "Unknown")
urban_levels <- c("Urban", "Rural", "Unknown")

# ---- helper functions ----

# Recode race/ethnicity to harmonized values and create EthRace
recode_ethrace <- function(df, race_col = "prefrace", eth_col = "prefethnicity") {
  df %>%
    mutate(
      !!sym(race_col) := case_when(
        .data[[race_col]] == "Black or African American" ~ "African American",
        .data[[race_col]] == "Native Hawaiian or Other Pacific Islander" ~ "Asian",
        .data[[race_col]] == "American Indian or Alaska Native" ~ "Native American",
        .data[[race_col]] == "some other race" ~ "Other",
        TRUE ~ .data[[race_col]]
      ),
      !!sym(eth_col) := case_when(
        .data[[eth_col]] == "Non-Hispanic" ~ "Non-His",
        .data[[eth_col]] == "Hispanic or Latino" ~ "His_x",
        TRUE ~ .data[[eth_col]]
      )
    ) %>%
    mutate(
      EthRace = paste(.data[[eth_col]], .data[[race_col]], sep = " "),
      EthRace = case_when(
        grepl("His_x", EthRace) ~ "Hispanic",
        TRUE ~ EthRace
      ),
      EthRace = case_when(
        EthRace %in% c("Non-His Unknown", "Unknown Other", "Unknown White",
                       "Unknown Black", "Unknown Unknown", "Unknown Asian",
                       "Unknown Native A", "Unknown Pacific I") ~ "Unknown",
        TRUE ~ EthRace
      ),
      EthRace = gsub("Non-His ", "", EthRace),
      EthRace = gsub("Unknown ", "", EthRace)
    )
}

# Standardize age groups
standardize_age <- function(df) {
  df %>%
    mutate(age_group = case_when(
      age_group %in% c("Under 1", "1 to 4") ~ "Under 5",
      TRUE ~ age_group
    ))
}

# Clean urban/rural & gender for display
standardize_demographics <- function(df) {
  df %>%
    mutate(
      prefurban = case_when(
        tolower(prefurban) == "urban" ~ "Urban",
        tolower(prefurban) == "rural" ~ "Rural",
        is.na(prefurban) ~ "Unknown",
        TRUE ~ prefurban
      ),
      prefgender = case_when(
        prefgender == "Other or Unknown" ~ "Unknown",
        TRUE ~ prefgender
      )
    )
}

# Create incidence map with hatching for low counts
make_state_map <- function(df, value_col, cut_breaks, cut_labels,
                           fill_palette, title, legend_title,
                           hatch_threshold = 5, hatch_data = NULL) {
  stopifnot("state" %in% names(df))
  df <- df %>% mutate(cut_val = cut(.data[[value_col]],
                                   breaks = cut_breaks,
                                   labels = cut_labels,
                                   include.lowest = TRUE))
  df$cut_val[df[[value_col]] == 1 & "1-10" %in% cut_labels] <- "1-10" # special case if needed
  df$hatch <- ifelse(df[[value_col]] < hatch_threshold, TRUE, FALSE)
  
  p <- plot_usmap(data = df, values = "cut_val") +
    scale_fill_manual(values = fill_palette, name = legend_title) +
    labs(title = title) +
    theme(panel.background = element_blank(), legend.position = "right") +
    guides(fill = guide_legend(title = legend_title)) +
    # add hatching for low counts
    geom_sf_pattern(
      data = df %>% filter(hatch == TRUE),
      aes(geometry = geom, pattern = hatch),
      fill = NA,
      pattern_angle = 45,
      pattern_density = 0.05,
      pattern_spacing = 0.02,
      color = "black",
      size = 0.01
    ) +
    scale_pattern_manual(values = c("TRUE" = "stripe"), name = "< threshold", labels = c("TRUE" = "")) +
    guides(pattern = guide_legend(title = "< threshold"))
  p
}

# Fit linear trend per state
fit_state_trends <- function(df, value_col = "estimate") {
  df_list <- split(df, df$prefstate)
  df_list <- map(df_list, ~ mutate(.x, Year = as.numeric(as.character(Year))))
  lm_list <- map(df_list, ~ {
    if (nrow(.x) >= 3 && all(.x[[value_col]] > 0, na.rm = TRUE)) {
      lm_formula <- reformulate("Year", response = value_col)
      lm(lm_formula, data = .x)
    } else NULL
  })
  res <- imap_df(lm_list, ~ {
    if (is.null(.x)) return(NULL)
    tidy(.x) %>%
      filter(term == "Year") %>%
      mutate(State = .y)
  })
  res
}

# Utility to build time-blocked incidence summaries
summarize_by_time_block <- function(df, year_range, weighted = FALSE) {
  low <- year_range[1]; high <- year_range[2]
  sub <- df %>%
    filter(as.numeric(as.character(Year)) >= low,
           as.numeric(as.character(Year)) <= high) %>%
    group_by(prefstate) %>%
    summarize(
      nFungalPts = if (weighted) sum(weighted_nPatient, na.rm = TRUE) else sum(nFungalPts, na.rm = TRUE),
      nALLpts = sum(Sample_Strata_Denom, na.rm = TRUE)
    ) %>%
    mutate(
      measure = 100000 * (nFungalPts / nALLpts)
    ) %>%
    rename(state = prefstate)
  sub
}
```

## **1. Data loading & initial cleaning**
```{r}
# Read raw data
crypto_data <- read_csv("") %>%
  rename("Year" = "YEAR") %>%
  recode_ethrace() %>%
  standardize_age() %>%
  standardize_demographics()

Weighted_denominators <- read_csv("") %>%
  mutate(Year = as.factor(Year))

Unweighted_denominators_month <- read_csv("") %>%
  rename("Year" = "date2_year", "Month" = "date2_month")

Unweighted_denominators_year <- read_csv("") %>%
  rename("Year" = "date2_year") %>%
  filter(Year >= 2013 & Year <= 2023) %>%
  recode_ethrace() %>%
  standardize_age() %>%
  standardize_demographics()

climate_regions_map <- read_csv("") %>%
  rename(prefstate = state_code)

# Factor conversions
crypto_data <- crypto_data %>%
  mutate(
    Year = as.factor(Year),
    EthRace = factor(EthRace, levels = ethrace_levels),
    age_group = factor(age_group, levels = c("Under 5", "5 to 9","10 to 14", "16 to 19","20 to 24", "25 to 29", "30 to 34", "35 to 39",
                                            "40 to 44", "45 to 49", "50 to 54", "55 to 59","60 to 64", "65 and Over", "Unknown")),
    prefgender = factor(prefgender, levels = gender_levels),
    prefurban = factor(prefurban, levels = urban_levels)
  )
```

## **2. Aggregations (grouped data)**
```{r}
# Yearly grouped
crypto_data_grouped_year <- crypto_data %>%
  group_by(Year, prefstate, prefurban, EthRace, prefgender, age_group,
           category_Immunocompetent, category_HIV_AIDS, category_Immunocompromised_Unspecified,
           category_Leukemia, category_Lymphoma, category_Medication_Suppression,
           category_Organ_Transplant) %>%
  summarize(nPatients = sum(nPatients), .groups = "drop") %>%
  mutate(Year = as.factor(Year))

# Unweighted denominators annual grouped
Unweighted_denominators_year_grouped <- Unweighted_denominators_year %>%
  group_by(Year, prefstate, prefurban, EthRace, prefgender, age_group) %>%
  summarize(nPatient = sum(nPatient), .groups = "drop") %>%
  mutate(Year = as.factor(Year))

```

## **3. Weighted Analysis**
```{r}
# Prepare weighted crypto data
weighted.crypto_data_comorbids <- crypto_data_grouped_year %>%
  group_by(Year, prefstate, EthRace, prefgender, age_group,
           category_Immunocompetent, category_HIV_AIDS,
           category_Immunocompromised_Unspecified, category_Leukemia,
           category_Lymphoma, category_Medication_Suppression,
           category_Organ_Transplant) %>%
  summarize(nPatient = sum(nPatients), .groups = "drop") %>%
  filter(
    EthRace != "Unknown", prefgender != "Other or Unknown",
    !is.na(prefstate), age_group != "not mapped", age_group != "Unknown"
  ) %>%
  mutate(Year = as.factor(Year))

weighted.crypto_data <- crypto_data_grouped_year %>%
  group_by(Year, prefstate, prefurban, EthRace, prefgender, age_group) %>%
  summarize(nPatient = sum(nPatients), .groups = "drop") %>%
  filter(
    EthRace != "Unknown", prefgender != "Other or Unknown",
    !is.na(prefstate), age_group != "not mapped", age_group != "Unknown"
  ) %>%
  mutate(Year = as.factor(Year))

# Join denominators
weighted.crypto_data_comorbids <- full_join(
  weighted.crypto_data_comorbids,
  Weighted_denominators,
  by = c("Year", "prefstate", "prefgender", "EthRace", "age_group")
) %>%
  filter(!is.na(Sample_Strata_Denom)) %>%
  mutate(nPatient = replace_na(nPatient, 0))

weighted.crypto_data_final_merged <- full_join(
  weighted.crypto_data,
  Weighted_denominators,
  by = c("Year", "prefstate", "prefgender", "EthRace", "age_group")
) %>%
  filter(!is.na(Sample_Strata_Denom)) %>%
  mutate(nPatient = replace_na(nPatient, 0),
         weighted_nPatient = nPatient * weights,
         Standard_Inc = weighted_nPatient / ACS_Strata_Denom)
```

## **4. Table 1: Demographics of fungal cohort**
```{r}
table_1_data <- crypto_data %>%
  filter(nPatients > 0) %>%
  uncount(nPatients) %>%
  standardize_demographics() %>%
  mutate(
    prefgender = factor(prefgender, levels = gender_levels),
    prefurban = factor(prefurban, levels = urban_levels),
    EthRace = factor(EthRace, levels = ethrace_levels),
    age_group = factor(age_group, levels = age_levels_table1)
  ) %>%
  rename(
    "Gender" = prefgender,
    "Patient Location" = prefurban,
    "Age Group" = age_group,
    "Ethnicity and Race" = EthRace
  )

my.vars <- c("Year", "Gender", "Patient Location", "Age Group", "Ethnicity and Race",
             "category_Immunocompetent", "category_HIV_AIDS", "category_Immunocompromised_Unspecified",
             "category_Leukemia", "category_Lymphoma", "category_Medication_Suppression",
             "category_Organ_Transplant")
my.fvars <- my.vars


formatted_table <- table_1_data %>%
  tbl_summary(
    include = c("Gender", "Patient Location", "Ethnicity and Race", "Age Group")
  ) %>%
  as_kable_extra(booktabs = TRUE) %>%
  kable_styling(font_size = 12, position = "center", latex_options = "hold_position") %>%
  kable_classic(full_width = FALSE, html_font = "Cambria")

formatted_table

```

## **5. Incidence over time (unweighted vs weighted)**
```{r}
# Prepare summary incidence data
standardized_weighted.crypto_data <- weighted.crypto_data %>%
  rename(nFungalPts = nPatient) %>%
  mutate(E_nFungalPts = weights*nFungalPts)

summary_data_py <- standardized_weighted.crypto_data %>%
  group_by(Year) %>%
  summarize(
    nFungalPts = sum(as.numeric(nFungalPts), na.rm = TRUE),
    E_nFungalPts = sum(as.numeric(E_nFungalPts), na.rm = TRUE),
    nALLpts = sum(as.numeric(Sample_Strata_Denom), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    estimate = 100000 * (nFungalPts / nALLpts),
    weighted_estimate = 100000 * (E_nFungalPts / nALLpts)
  )

# Poisson regression for trend
mod_unweighted <- glm(nFungalPts ~ -1 + offset(log(nALLpts)) + as.factor(Year),
                      data = summary_data_py, family = poisson())
mod_weighted <- glm(E_nFungalPts ~ -1 + offset(log(nALLpts)) + as.factor(Year),
                    data = summary_data_py, family = poisson())

# Extract estimates for ribbons
mod_1_output <- tidy(mod_unweighted, exponentiate = TRUE, conf.int = TRUE) %>%
  select(-term, -statistic, -p.value, -std.error) %>%
  rename(U_estimate = estimate, U_conf.low = conf.low, U_conf.high = conf.high) %>%
  mutate(across(everything(), ~ . * 100000))
mod_2_output <- tidy(mod_weighted, exponentiate = TRUE, conf.int = TRUE) %>%
  select(-term, -statistic, -p.value, -std.error) %>%
  rename(W_estimate = estimate, W_conf.low = conf.low, W_conf.high = conf.high) %>%
  mutate(across(everything(), ~ . * 100000))

summary_data_py <- bind_cols(summary_data_py, mod_1_output, mod_2_output)

# Plot
plot_IR_py <- summary_data_py %>%
  ggplot(aes(x = as.factor(Year), group = 1)) +
  geom_point(aes(y = as.numeric(U_estimate), color = "Sample Incidence")) +
  geom_line(aes(y = as.numeric(U_estimate), color = "Sample Incidence")) +
  geom_ribbon(aes(ymin = U_conf.low, ymax = U_conf.high, fill = "Sample Incidence"),
              alpha = 0.2, linetype = 2) +
  geom_point(aes(y = as.numeric(W_estimate), color = "Population Incidence")) +
  geom_line(aes(y = as.numeric(W_estimate), color = "Population Incidence")) +
  geom_ribbon(aes(ymin = W_conf.low, ymax = W_conf.high, fill = "Population Incidence"),
              alpha = 0.2, linetype = 2) +
  scale_color_manual(
    values = c("Population Incidence" = "#d95f02", "Sample Incidence" = "#1b9e77"),
    name = "Incidence Estimates"
  ) +
  scale_fill_manual(
    values = c("Population Incidence" = "#d95f02", "Sample Incidence" = "#1b9e77"),
    name = "Incidence Estimates"
  ) +
  labs(
    x = "Year",
    y = "Incidence Rate (per 100,000)",
    title = "Sample and Population-Level Incidence Rates of Cryptococcosis, 2013–2023"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")

print(plot_IR_py)
if (params$save_plots) ggsave("plot_IR_py.jpeg", plot = plot_IR_py, width = 6, height = 4, dpi = 300)

```

## **6. State-level incidence maps**
```{r}
# Prepare summary for full period (unweighted)
summary_data_state <- crypto_data %>%
  group_by(prefstate) %>%
  summarize(nFungalPts = sum(nPatients), nALLpts = sum(nPatient, na.rm = TRUE), .groups = "drop") %>%
  mutate(Incidence = 1000000 * (nFungalPts / nALLpts)) %>%
  rename(state = prefstate)

# Merge boundary and plot
boundary_data <- us_map() %>%
  select(-full) %>%
  rename(state = abbr) %>%
  st_as_sf(crs = 4326)

plot_full <- make_state_map(
  df = left_join(summary_data_state, boundary_data, by = "state") %>% filter(!is.na(fips)),
  value_col = "Incidence",
  cut_breaks = c(0, 0.00000000000001, 2.5, 5, 7.5, 10, 15, 20, 25, Inf),
  cut_labels = c("0", "<2.5", "2.5-5", "5-7.5", "7.5-10", "10-15", "15-20", "20-25", "25+"),
  fill_palette = c("0" = "lightgrey", "<2.5" = "#ffeda0", "2.5-5" = "#fed976",
                   "5-7.5" = "#feb24c", "7.5-10" = "#fd8d3c",
                   "10-15" = "#fc4e2a", "15-20" = "#e31a1c", "20-25" = "#bd0026", "25+" = "#800026"),
  title = "Cryptococcosis Incidence Per State, 2013-2023",
  legend_title = "Incidence per 1,000,000"
)
print(plot_full)
if (params$save_plots) ggsave("incidence_full_period.jpeg", plot = plot_full, dpi = 300, width = 10, height = 6)

# Time blocks example loop
time_blocks <- list(
  `2013_2015` = c(2013, 2015),
  `2016_2018` = c(2016, 2018),
  `2019_2021` = c(2019, 2021),
  `2022_2023` = c(2022, 2023)
)

for (name in names(time_blocks)) {
  yr <- time_blocks[[name]]
  block_df <- summarize_by_time_block(weighted.crypto_data_final_merged, yr, weighted = FALSE)
  block_map_df <- left_join(block_df, boundary_data, by = c("state" = "state")) %>%
    filter(!is.na(fips))
  p <- make_state_map(
    df = block_map_df,
    value_col = "measure",
    cut_breaks = c(0, 0.00000000000001, 2.5, 5, 7.5, 10, 15, 20, 25, Inf),
    cut_labels = c("0", "<2.5", "2.5-5", "5-7.5", "7.5-10", "10-15", "15-20", "20-25", "25+"),
    fill_palette = c("0" = "lightgrey", "<2.5" = "#ffeda0", "2.5-5" = "#fed976",
                     "5-7.5" = "#feb24c", "7.5-10" = "#fd8d3c",
                     "10-15" = "#fc4e2a", "15-20" = "#e31a1c", "20-25" = "#bd0026", "25+" = "#800026"),
    title = paste0("Incidence: ", gsub("_", "-", name)),
    legend_title = "Incidence per 100,000"
  )
  print(p)
  if (params$save_plots) ggsave(paste0("incidence_", name, ".jpeg"), plot = p, dpi = 300, width = 10, height = 6)
}
```

## **7. Trend analysis by state**
```{r}
# Prepare state-year incidence for weighted data
state_year <- weighted.crypto_data_final_merged %>%
  group_by(Year, prefstate) %>%
  summarize(
    nPatient = sum(weighted_nPatient, na.rm = TRUE),
    Sample_Strata_Denom = sum(Sample_Strata_Denom, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    estimate = 100000 * (nPatient / Sample_Strata_Denom),
    prefstate = factor(prefstate)
  ) %>%
  filter(prefstate != "DC") # optional exclusion

trend_results <- fit_state_trends(state_year, value_col = "estimate")
print(trend_results %>% arrange(p.value))

```

## **8. Group-specific incidence trends**
### **8.1 Ethnicity / Race trends**
```{r ethnicity-trends}
# Weighted incidence by EthRace over time
ethrace_trends <- weighted.crypto_data_final_merged %>%
  filter(!grepl("Unknown", EthRace)) %>%
  group_by(Year, EthRace) %>%
  summarize(
    weighted_nPatient = sum(weighted_nPatient, na.rm = TRUE),
    Sample_Strata_Denom = sum(Sample_Strata_Denom, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(estimate = 100000 * (weighted_nPatient / Sample_Strata_Denom),
         EthRace = factor(EthRace, levels = ethrace_levels))

ggplot(ethrace_trends, aes(x = as.numeric(Year), y = estimate, group = EthRace, color = EthRace)) +
  geom_point(size = 0.75) +
  geom_line() +
  facet_wrap(~EthRace, nrow = 1, strip.position = "bottom") +
  scale_x_continuous(breaks = seq(2014, 2023, by = 3)) +
  labs(
    title = "Cryptococcosis Incidence by Ethnicity and Race, 2013-2023",
    x = "Year",
    y = "Incidence per 100,000",
    color = "Ethnicity and Race"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
```

### **8.2 Age trends**
```{r}
age_trends <- weighted.crypto_data_final_merged %>%
  filter(!grepl("Unknown", age_group)) %>%
  group_by(Year, age_group) %>%
  summarize(
    weighted_nPatient = sum(weighted_nPatient, na.rm = TRUE),
    Sample_Strata_Denom = sum(Sample_Strata_Denom, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    estimate = 100000 * (weighted_nPatient / Sample_Strata_Denom),
    age_group = factor(age_group, levels = c("Under 1", "1 to 4", "5 to 9", "10 to 14", "15 to 19",
                                             "20 to 24", "25 to 29", "30 to 34", "35 to 39",
                                             "40 to 44", "45 to 49", "50 to 54", "55 to 59",
                                             "60 to 64", "65 and Over"))
  )

ggplot(age_trends, aes(x = as.numeric(Year), y = estimate, group = age_group, color = age_group)) +
  geom_point(size = 0.6) +
  geom_line() +
  facet_wrap(~age_group, nrow = 1, strip.position = "bottom") +
  scale_x_continuous(breaks = seq(2014, 2023, by = 3)) +
  labs(
    title = "Cryptococcosis Incidence by Age Group, 2013-2023",
    x = "Year",
    y = "Incidence per 100,000",
    color = "Age Group"
  ) +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## **9. Multivariable regression with robust standard errors & adjusted IRR plots**
### **9.1 Helper for quasi-Poisson with robust SEs**
```{r}
library(sandwich)
library(broom)

# Fit quasi-Poisson and return exponentiated coefficients with robust CIs
fit_qp <- function(formula, data, offset_var, weight_var = "weights") {
  mod <- glm(formula, data = data,
             family = quasipoisson(link = "log"),
             offset = log(data[[offset_var]]),
             weights = data[[weight_var]])
  vcov_robust <- vcovHC(mod, type = "HC0")
  tidy_mod <- tidy(mod, exponentiate = TRUE, conf.int = TRUE, vcov. = vcov_robust)
  tidy_mod
}

```

### **9.2 Prepare modeling dataset**
```{r}
# Build immuno_status from hot-encoded comorbidities
model_df <- weighted.crypto_data_final_merged %>%
  mutate(
    immuno_status = case_when(
      (category_HIV_AIDS + category_Immunocompromised_Unspecified + category_Leukemia +
         category_Lymphoma + category_Medication_Suppression + category_Organ_Transplant) > 1 ~ "Multiple",
      category_HIV_AIDS == 1 ~ "HIV/AIDS",
      category_Immunocompromised_Unspecified == 1 ~ "Immunocompromised_Unspecified",
      category_Leukemia == 1 ~ "Leukemia",
      category_Lymphoma == 1 ~ "Lymphoma",
      category_Medication_Suppression == 1 ~ "Medication_Suppression",
      category_Organ_Transplant == 1 ~ "Organ_Transplant",
      category_Immunocompetent == 1 ~ "Immunocompetent",
      TRUE ~ "Immunocompetent"
    ),
    # clean up levels / references
    EthRace = relevel(factor(EthRace, levels = ethrace_levels), ref = "White"),
    prefgender = relevel(factor(prefgender, levels = gender_levels), ref = "Female"),
    prefurban = relevel(factor(prefurban, levels = urban_levels), ref = "Rural"),
    age_group = relevel(factor(age_group), ref = "20 to 24"),
    prefstate = relevel(factor(prefstate), ref = "IN"),
    immuno_status = relevel(factor(immuno_status), ref = "Immunocompetent")
  ) %>%
  # join climate region if needed
  left_join(climate_regions_map, by = c("prefstate" = "prefstate")) %>%
  mutate(climate_regions = if_else(is.na(climate_regions), "Unknown", climate_regions)) %>%
  mutate(climate_regions = relevel(factor(climate_regions), ref = "NORTHEN ROCKIES AND PLAINS (WEST NORTH CENTRAL)")) %>%
  # ensure numeric offset
  rename(nFungalPts = nPatient,
         nALLpts = Sample_Strata_Denom)

```

### **9.3 Full adjusted model**
```{r}
full_formula <- as.formula("nFungalPts ~ as.factor(Year) + EthRace + prefgender + prefurban + age_group + prefstate + immuno_status + climate_regions")
full_mod <- fit_qp(full_formula, data = model_df, offset_var = "nALLpts", weight_var = "weights")
full_mod %>% arrange(p.value) %>% head(10)
```

### **9.4 Plotting function for adjusted IRRs**
```{r}
plot_irr <- function(tidy_df, pattern, ref_label, title, x_label = "Adjusted Incidence Rate Ratio", ylim = c(0, NA)) {
  df <- tidy_df %>%
    filter(grepl(pattern, term)) %>%
    mutate(
      term_clean = gsub(pattern, "", term),
      term_clean = str_replace_all(term_clean, "^", ""), # extra cleanup if needed
      term_clean = str_trim(term_clean)
    ) %>%
    # add reference row if missing
    bind_rows(
      tibble(
        term = NA_character_,
        estimate = 1,
        conf.low = 1,
        conf.high = 1,
        term_clean = ref_label,
        p.value = NA,
        std.error = NA,
        statistic = NA
      )
    ) %>%
    distinct(term_clean, .keep_all = TRUE) %>%
    arrange(estimate) %>%
    mutate(label = sprintf("%.2f", estimate))
  
  ggplot(df, aes(x = reorder(term_clean, estimate), y = estimate)) +
    geom_point() +
    geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
    geom_hline(yintercept = 1, linetype = "dashed") +
    coord_flip() +
    labs(title = title, x = NULL, y = x_label, caption = "Reference = baseline") +
    theme_minimal() +
    scale_y_continuous(expand = expansion(mult = c(0, .1))) +
    geom_text(aes(label = label), vjust = -0.5, size = 3)
}
```

### **9.5 Adjusted IRR plots for key exposures**
```{r}
# Ethnicity / Race
p_ethrace <- plot_irr(full_mod, pattern = "EthRace", ref_label = "NH White",
                      title = "Adjusted IRRs for Cryptococcosis Diagnosis by Ethnicity and Race")
print(p_ethrace)
if (params$save_plots) ggsave("adj_IRR_ethrace.jpeg", plot = p_ethrace, width = 8, height = 5)

# Age Group
p_agegroup <- plot_irr(full_mod, pattern = "age_group", ref_label = "20 to 24",
                       title = "Adjusted IRRs for Cryptococcosis Diagnosis by Age Group")
print(p_agegroup)
if (params$save_plots) ggsave("adj_IRR_agegroup.jpeg", plot = p_agegroup, width = 8, height = 5)

# Gender
p_gender <- plot_irr(full_mod, pattern = "prefgender", ref_label = "Female",
                     title = "Adjusted IRRs for Cryptococcosis Diagnosis by Gender")
print(p_gender)
if (params$save_plots) ggsave("adj_IRR_gender.jpeg", plot = p_gender, width = 6, height = 4)

# Urban/Rural
# Note: variable name is prefurban
p_urban <- plot_irr(full_mod, pattern = "prefurban", ref_label = "Rural",
                    title = "Adjusted IRRs for Cryptococcosis Diagnosis by Patient Location")
print(p_urban)
if (params$save_plots) ggsave("adj_IRR_urban_rural.jpeg", plot = p_urban, width = 6, height = 4)

# Comorbidities (immuno_status)
p_comorbid <- plot_irr(full_mod, pattern = "immuno_status", ref_label = "Immunocompetent",
                       title = "Adjusted IRRs for Cryptococcosis Diagnosis by Comorbidity Status")
print(p_comorbid)
if (params$save_plots) ggsave("adj_IRR_comorbid.jpeg", plot = p_comorbid, width = 8, height = 5)

# Climate regions
p_climate <- plot_irr(full_mod, pattern = "climate_regions", ref_label = "NORTHEN ROCKIES AND PLAINS (WEST NORTH CENTRAL)",
                      title = "Adjusted IRRs for Cryptococcosis Diagnosis by Climate Region")
print(p_climate)
if (params$save_plots) ggsave("adj_IRR_climate_regions.jpeg", plot = p_climate, width = 8, height = 6)
```

## **10. Spatial cluster analysis (LISA)**
### **10.1 Initialize Spatial Dataset**
```{r}
library(rgeoda)

# Prepare state-level weighted incidence (population incidence)
cluster_input <- weighted.crypto_data_final_merged %>%
  group_by(prefstate) %>%
  summarize(
    weighted_nPatient = sum(weighted_nPatient, na.rm = TRUE),
    Sample_Strata_Denom = sum(Sample_Strata_Denom, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Pop_Pt_Proportion_100000 = 100000 * (weighted_nPatient / Sample_Strata_Denom),
    state = prefstate
  )

# Spatial boundaries
us_states_sf <- us_map("states") %>%
  select(full, abbr, geom) %>%
  rename(state = abbr) %>%
  st_as_sf(crs = 4326) %>%
  group_by(state) %>%
  summarize(geometry = st_combine(geom), .groups = "drop") %>%
  st_cast("MULTIPOLYGON")

# Merge with incidence
cluster_sf <- left_join(us_states_sf, cluster_input, by = c("state" = "state")) %>%
  filter(!is.na(Pop_Pt_Proportion_100000))

# Build spatial weights and local Moran's I
w <- queen_weights(cluster_sf)
lisa <- local_moran(w, cluster_sf["Pop_Pt_Proportion_100000"])

# Annotate clusters
cluster_sf$lisa_cluster <- factor(lisa_clusters(lisa),
                                  levels = 0:4,
                                  labels = c("Not Significant", "High-High", "Low-Low", "Low-High", "High-Low"))
cluster_sf$lisa_p_values <- lisa$p_vals
cluster_sf$hatch <- ifelse(cluster_sf$weighted_nPatient < 5, TRUE, FALSE)
```

### **10.2 Plot Spatial Clusters**
```{r}
# LISA cluster map
lisa_palette <- c("Not Significant" = "lightgray",
                  "Low-Low" = "#1b9e77",
                  "High-Low" = "#fc9272",
                  "Low-High" = "#66c2a5",
                  "High-High" = "#d95f02")

ggplot() +
  geom_sf(data = cluster_sf, aes(fill = lisa_cluster), color = NA) +
  scale_fill_manual(values = lisa_palette, name = "LISA Cluster") +
  # hatching for low counts if desired
  geom_sf_pattern(
    data = cluster_sf %>% filter(hatch == TRUE),
    aes(geometry = geometry, pattern = hatch),
    fill = NA,
    pattern_angle = 45,
    pattern_density = 0.05,
    pattern_spacing = 0.02,
    color = "black",
    size = 0.1
  ) +
  scale_pattern_manual(values = c("TRUE" = "stripe"), guide = "none") +
  labs(
    title = "LISA Cluster Map of State-Level Weighted Cryptococcosis Incidence",
    subtitle = "Population incidence per 100,000 (2013–2023)",
    caption = "Clusters: High-High, Low-Low, spatial outliers"
  ) +
  theme_minimal() +
  theme(legend.position = "right") 

if (params$save_plots) ggsave("lisa_cluster_map.jpeg", width = 10, height = 6, dpi = 300)
```

### **10.3 Extract Significant Clusters**
```{r}
significant_clusters <- cluster_sf %>%
  filter(lisa_cluster != "Not Significant") %>%
  select(state, lisa_cluster, lisa_p_values, Pop_Pt_Proportion_100000)
print(significant_clusters)
```

